### 2016-09-09 Chapters/ResponseTheory.tex
Finished response theory chapter. One section in appendix missing

The variational setting introduced in Section \ref{sec:variational},
offers a series of advantages also in the formulation of response
theory. As already noted, a classical counterpart to the
Hellmann--Feynman theorem holds:
\begin{equation}\label{eq:classical-HF}
  \deriv{U_\mathrm{pol}}{F} = \pderiv{U_\mathrm{pol}}{F}
  + \pderiv{U_\mathrm{pol}}{\s}\pderiv{\s}{F}
  = \pderiv{U_\mathrm{pol}}{F},
\end{equation}
which forms the basis for our formal development in
\paper{V}.\autocite{pcm-openrsp}
There we show how the variational polarization functional can
successfully achieve a transparent derivation of arbitrary order
response functions, when the fixed-cavity approximation for the
\acrshort*{PCM} is assumed.
Here we report a summary of our findings in \paper{V} employing the
classical functional in supermatrix form in Eq. \eqref{eq:supermatrix-functional}.
In the notation of \citeauthor{Thorvaldsen2008-sg}, we introduce the
generalized free energy:
\begin{equation}\label{eq:generalized-KS-energy}
  \begin{aligned}
 \tilde{\mathcal{G}}
 &=
 \tilde{\mathcal{E}}
 + \frac{1}{2}{}^t\tilde{\p}\V\tilde{\p} +
 {}^t\tilde{\p}\Tr(\tilde{\mathbfup{s}}\pertmat{D}) \\
 &\eqtr
 \left[\pertmat{h} + \pertmat{V}^{t} +
 \frac{1}{2}\pertmat{G}^{\gamma}(\pertmat{D}) -
 \frac{\mathrm{i}}{2}\pertmat{T}\right]\pertmat{D}
 + \tilde{E}_\mathrm{xc}[\tilde{\rho}(\pertmat{D})] + h_\mathrm{nuc}
 + \frac{1}{2}{}^t\tilde{\p}\V\tilde{\p}
 + {}^t\tilde{\p}\tilde{\mathbfup{s}}\pertmat{D},
 \end{aligned}
\end{equation}
whose perturbation-strength derivatives at zero perturbation strength
will yield the desired response functions.

Due to the implicit time-dependence of $\pertmat{D}$ and
$\tilde{\sigma}$, higher-order
derivatives of the KS generalized energy will require application of the
chain rule. The ${mn, abc\ldots}$ superscript describes how and to which
extent the chain rule was applied for a given term, \ie the number of explicit
differentiations with respect to the variational densities, so that
\begin{equation}\label{eq:differential}
  \mathcal{G}^{mn, abc}
  =
  \frac{\partial^{m+n+3}\mathcal{G}}{\partial(\mat{D}^T)^m\partial\p^n\partial\epsilon_a\partial\epsilon_b\partial\epsilon_c}
  =
    \mathcal{E}^{m, abc} +
  \frac{\partial^{m+n+3}U_\mathrm{pol}}{\partial(\mat{D}^T)^m\partial\p^n\partial\epsilon_a\partial\epsilon_b\partial\epsilon_c} \text{.}
\end{equation}
In this notation, the index $m$ denotes the order of differentiation
with respect to the density matrix $\mat{D}$, while the index $n$ symbolizes
the order of differentiation with respect to the \acs{ASC} density $\sigma$.
Differentiation with respect to the density matrix will result in a
$2m$-rank tensor, while differentiation with respect to the
\acs{ASC} density will result in a function of the continuous cavity
index $\vect{s}$. For higher-order properties, mixed terms involving both density
matrix and \acs{ASC} density differentiation may generally occur.
In the \emph{fixed-cavity approximation}, the cavity is kept frozen at a
given molecular geometry.\autocite{Cammi1994-qj}
Under this simplifying assumption, only the linear interaction term in
the polarization functional eq.~\ref{eq:PCM-functional} will be affected
by the movements of the nuclei \emph{via} the dependence of basis functions on
the molecular geometry.
Its perturbation-strength derivative will then be
\begin{equation}\label{eq:PCM-derivative}
 \deriv{}{\epsilon_a}\left\lbrace
 U_\mathrm{pol}
 \right\rbrace_T
 \eqavetr
 {}^t\tilde{\p}\tilde{\mathbfup{s}}^a\pertmat{D} \text{,}
\end{equation}
where the second term only involves derivatives of the electrostatic potential integrals.
We remark that, under the fixed-cavity approximation, both the density
matrix -- $m$ -- and \acs{ASC} density -- $n$ -- differentiation indices
in eq.~\ref{eq:differential} can only assume the values 0 or 1 for the
$\frac{\partial^{m+n+3}U_\mathrm{pol}}{\partial(\mat{D}^T)^m\partial\p^n\partial\epsilon_a\partial\epsilon_b\partial\epsilon_c}$
term.
By construction, the density matrix dependence in the
polarization functional is at most linear, while,
by virtue of the classical Hellmann--Feynman theorem,
eq.~\ref{eq:classical-HF}, the \acs{ASC} variational density will also appear
at most \emph{linearly} in $\tilde{\mathcal{G}}^{00,a}$.
The free energy term perturbation strength derivative is given as
\begin{equation}\label{eq:tilde-G00-a}
  \begin{aligned}
    \tilde{\mathcal{G}}^{00,a}(\tilde{\mat{D}}, \tilde{\sigma}, t)
    &= \tilde{\mathcal{E}}^{0,a} +
    {}^t\tilde{\p}\Tr({\tilde{\mathbfup{s}}^a\pertmat{D}}) \\
  &\eqtr
  (\pertmat{h}^a + \pertmat{V}^{t,a} + \frac{1}{2}\pertmat{G}^{\gamma, a}(\pertmat{D}) + \pertmat{F}^{\Omega_a}_\mathrm{xc}
  -\frac{\mathrm{i}}{2}\pertmat{T}^a)\pertmat{D}
  + h_\mathrm{nuc}^a + {}^t\tilde{\p}\tilde{\mathbfup{s}}^a\pertmat{D}
  \end{aligned}
\end{equation}
where the matrix $\pertmat{F}^{\Omega_a}_\mathrm{xc}$ denotes the
the functional derivative matrix defined in terms of the perturbed
overlap distributions $\tilde{\Omega}^a$.

Response functions can then be obtained by straightforward
differentiation with respect to further perturbations and subsequent
evaluation at zero perturbation strength, so that
\begin{subequations}
  \begin{align}
    L^a &\eqavetr
    \mathcal{G}^{00,a} - \mat{S}^a\mat{W} \label{eq:first-order} \\
    L^{ab} &\eqavetr \mathcal{G}^{00,ab}
    + \mathcal{G}^{10,a}\mat{D}^{b}
    + \mathcal{G}^{01,a}\p^{b}
    - \mat{S}^{ab}\mat{W}
    - \mat{S}^{a}\mat{W}^{b} \label{eq:second-order} \\
    L^{abc} &\eqavetr
      \mathcal{G}^{00,abc}
    + \mathcal{G}^{10,ac}\mat{D}^{b}
    + \mathcal{G}^{10,ab}\mat{D}^{c}
    + \mathcal{G}^{20,a}\mat{D}^{b}\mat{D}^{c} \nonumber \\
    &+ \mathcal{G}^{10,a}\mat{D}^{bc}
    + \mathcal{G}^{11,a}\mat{D}^{b}\p^c \label{eq:third-order} \\
    &+ \mathcal{G}^{01,ac}\p^{b}
    + \mathcal{G}^{01,ab}\p^{c}
    + \mathcal{G}^{01,a}\p^{bc}
    + \mathcal{G}^{11,a}\p^{b}\mat{D}^c \nonumber \\
    &- \mat{S}^{abc}\mat{W}
    - \mat{S}^{ab}\mat{W}^{c}
    - \mat{S}^{ac}\mat{W}^{b}
    - \mat{S}^{a}\mat{W}^{bc} \nonumber
  \end{align}
\end{subequations}
and similarly for higher-order response functions.

The governing equations for the perturbed \acs{ASC} densities are
obtained in complete analogy with the handling of pertubed density
matrices outlined above.
We introduce a decomposition of the \acs{ASC} density into frequency
components into the perturbation-strength derivative of
Eq. \eqref{eq:polarization-stat}, so that
\begin{equation}\label{eq:PCM-differentiated}
  \V\p_\omega^{b_N} + \Tr \mathbfup{s}\mat{D}_\omega^{b_N} =
  \mathbb{S}^{(n-1)}_\omega
  \eqtr - \left(\mat{\varphi}\mat{D}\right)_{\omega, n-1}^{b_N} \text{.}
\end{equation}
$\mathbb{S}^{(n-1)}_\omega$ contains all terms that depend on lower order density matrices and
differentiated source integrals.
The term $\mathbb{S}^{(n-1)}_\omega$ always contains at least a first derivative of the
electrostatic potential integrals and is thus zero if the basis set is
independent of the perturbation tuple under consideration.
We now introduce a partition of the frequency components of the
polarization $\p_\omega^{b_N}$ into homogeneous (H) and particular (P)
terms:
\begin{equation}\label{eq:polarization-partition}
  \p_\omega^{b_N} = \p_\mathrm{P}^{b_N} + \p_\mathrm{H}^{b_N} \text{,}
\end{equation}
a similar partition exists for the density matrix and we thus get
the following system of equations:
\begin{subequations}
  \begin{align}
    \V\p_\mathrm{H}^{b_N} &+ \Tr \mathbfup{s}\Dh{b_N} = 0
    \label{eq:homo-polarization-eq}\\
    \V\p_\mathrm{P}^{b_N} &+ \Tr \mathbfup{s}\Dp{b_N}
    = \mathbb{S}^{(n-1)}_\omega \text{.}
 \label{eq:part-polarization-eq}
  \end{align}
\end{subequations}
We note that the particular term of the polarization is nonzero if and
only if the basis set depends on the external perturbation.

We finally turn our attention to the \acs{TDSCF} equation.
The perturbation-strength differentiated generalized \acs{KS} matrix is
first separated into its frequency components $\mathbfcal{F}^{b_N}_\omega$.
The H/P partition introduced for the variational densities will induce a
similar partition into these frequency components:
\begin{equation}\label{eq:generalized-Fock-partition}
  \mathbfcal{F}_\omega^{b_N} =
  \mat{G}^{\mathrm{KS}}(\Dh{b_N}) + \ASCh{b_N}\mat{\esp}
  + \breve{\mathbfcal{F}}_\omega^{b_N} \text{.}
\end{equation}
The two-electron and \acs{XC} contributions depending on the homogeneous
perturbed density matrix have been collected into the
$\mat{G}^\mathrm{KS}(\Dh{b_N})$ matrix, while all other contributions
are collected into $\breve{\mathbfcal{F}}_\omega^{b_N}$, and we
refer to \noparcite[refs.][]{Thorvaldsen2008-sg} and
\noparcite[][]{Ringholm2014-gx} for a more detailed treatment of this.
The parametrization of the homogeneous part of the perturbed density
matrix can be exploited to conveniently reformulate the perturbed
\acs{TDSCF} equation, so that
\begin{equation}\label{eq:response-equation}
 \left[\genHessian - \omega_{b_N}\genMetric \right]\rspParam{b_N} = \RHS{b_N}
\end{equation}
where the generalized Hessian $\genHessian$ and metric $\genMetric$
matrices were introduced and are defined by their transformations on the
response parameters $\rspParam{b_N}$:\autocite{Larsen2000-hj,
Kjaergaard2008-hy}
\begin{align}
  \genHessian\rspParam{b_N} &=
  \mat{G}^\mathrm{KS}([\rspParam{b_N}, \mat{D}]_{\mat{S}})\mat{D}\mat{S}
  -\mat{S}\mat{D}\mat{G}^\mathrm{KS}([\rspParam{b_N}, \mat{D}]_{\mat{S}})
+\mat{F} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{S}
-\mat{S} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{F}
+ \ASCh{b_N}\mat{\esp}\mat{D}\mat{S}
-\mat{S}\mat{D}\mat{\esp}\ASCh{b_N}
\label{eq:E2-lintra}
\\
  \genMetric\rspParam{b_N} &= \mat{S} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{S} \text{.}
  \label{eq:S2-lintra}
\end{align}
The generalized Hessian matrix $\genHessian$ includes two types of
solvent term. \emph{Implicit} terms are those included in the
zeroth-order Fock matrix, $\mat{F}$.
The last two terms in Eq. \ref{eq:E2-lintra} are \emph{explicit} terms,
as they explicitly involve the $N$-th order homogenous \acs{ASC}
variational density.
The theoretical treatment of frequency-dependent properties in solution
within the \acs{PCM} requires adoption of a nonequilibrium response
framework.\autocite{Cammi1999-rb, Tomasi2005-vm}
The explicit \acs{PCM} terms in Eq. \ref{eq:E2-lintra} are then
evaluated using the optical permittivity $\dield$ of the solvent
instead of the static permittivity $\diels$ used to compute the
implicit contributions.
In contrast to $\genHessian$, the generalized metric matrix --
$\genMetric$ -- is unchanged.
The \ac{RHS} in the response equation only includes terms that
depend on particular contributions up to the desired order or lower-order perturbed density
matrices:
\begin{equation}\label{eq:RHS}
  \RHS{b_N} =
  \left[\left( \tilde{mathbfcal{F}}
-
\frac{\mathrm{i}}{2}\pertmat{S}\deriv{}{t}\right)\pertmat{D}\pertmat{S}\right]^{\ominus,
b_N}_\mathrm{P}
\end{equation}

Solution strategy for response equations~\autocite{Saad2003-oa,
Saad2011-gm, Kauczor2011-rd, Malmqvist2013-vw}

>>>>

The variational setting introduced in Section \ref{sec:variational},
offers a series of advantages also in the formulation of response
theory. As already noted, a classical counterpart to the
Hellmann--Feynman theorem holds:
\begin{equation}\label{eq:classical-HF}
  \deriv{U_\mathrm{pol}}{F} = \pderiv{U_\mathrm{pol}}{F}
  + \pderiv{U_\mathrm{pol}}{\s}\pderiv{\s}{F}
  = \pderiv{U_\mathrm{pol}}{F},
\end{equation}
which forms the basis for our formal development in
\paper{V}.\autocite{pcm-openrsp}
There we show how the variational polarization functional can
successfully achieve a transparent derivation of arbitrary order
response functions, when the fixed-cavity approximation for the
\acrshort*{PCM} is assumed.
Here we report a summary of our findings in \paper{V} employing the
supermatrix form of the classical functional as given in Eq.
\eqref{eq:supermatrix-functional}.
In the notation of \citeauthor{Thorvaldsen2008-sg}, we introduce the
generalized free energy:
\begin{equation}\label{eq:generalized-KS-energy}
  \begin{aligned}
 \tilde{\mathcal{G}}
 &=
 \tilde{\mathcal{E}}
 + \frac{1}{2}{}^t\tilde{\p}\V\tilde{\p} +
 {}^t\tilde{\p}\Tr(\tilde{\mathbfup{s}}\pertmat{D}) \\
 &\eqtr
 \left[\pertmat{h} + \pertmat{V}^{t} +
 \frac{1}{2}\pertmat{G}^{\gamma}(\pertmat{D}) -
 \frac{\mathrm{i}}{2}\pertmat{T}\right]\pertmat{D}
 + \tilde{E}_\mathrm{xc}[\tilde{\rho}(\pertmat{D})] + h_\mathrm{nuc}
 + \frac{1}{2}{}^t\tilde{\p}\V\tilde{\p}
 + {}^t\tilde{\p}\tilde{\mathbfup{s}}\pertmat{D},
 \end{aligned}
\end{equation}
whose perturbation-strength derivatives at zero perturbation strength
will yield the desired response functions.

To keep track of the number of chain rule applications,
we will denote higher-order derivatives of the \acrshort*{KS}
generalized free energy with the composite ${mn, abc\ldots}$
superscript:
\begin{equation}\label{eq:differential}
  \mathcal{G}^{mn, abc}
  =
  \frac{\partial^{m+n+3}\mathcal{G}}{\partial(\mat{D}^T)^m\partial\p^n\partial\epsilon_a\partial\epsilon_b\partial\epsilon_c}
  =
    \mathcal{E}^{m, abc} +
  \frac{\partial^{m+n+3}U_\mathrm{pol}}{\partial(\mat{D}^T)^m\partial\p^n\partial\epsilon_a\partial\epsilon_b\partial\epsilon_c} \text{.}
\end{equation}
In this notation, the $abc\ldots$ indices denote the perturbations with
respect to which differentiation occurs. The index $m$ denotes the order
of differentiation with respect to the density matrix. Eventually, $n$
eventually denotes the order of differentiation with respect to the
polarization.
The classical interaction kernel $\V$ might, in the most general
scenario, depend on the molecular geometry.
In \paper{V} and in the following we ignore this possibility and work
under the simplifying assumption that $\V$ is independent of the nuclear
positions.
This goes under the name of \emph{fixed-cavity approximation} in the
context of the \acrshort*{PCM}.\autocite{Cammi1994-qj}
Under this simplifying assumption, the perturbation-strength derivative
of the classical functional will only include terms stemming from the
linear interaction term:
\begin{equation}\label{eq:polarization-derivative}
 \deriv{}{\epsilon_a}\left\lbrace
 U_\mathrm{pol}
 \right\rbrace_T
 \eqavetr
 {}^t\tilde{\p}\tilde{\mathbfup{s}}^a\pertmat{D} \text{,}
\end{equation}
where the second term only involves derivatives of the source integrals.
Moreover, in the general expression for the perturbation-strength
derivative of the polarization functional in Eq. \eqref{eq:differential},
the $m$ and $n$ indices can only assume the values 0 or 1.
In fact, the density matrix dependence is at most linear in the
polarization functional, while the classical Hellmann--Feynman theorem
ensures that the polarization appears at most linearly in $\tilde{\mathcal{G}}^{00,a}$:
\begin{equation}\label{eq:tilde-G00-a}
  \begin{aligned}
    \tilde{\mathcal{G}}^{00,a}(\tilde{\mat{D}}, \tilde{\sigma}, t)
    &= \tilde{\mathcal{E}}^{0,a} +
    {}^t\tilde{\p}\Tr({\tilde{\mathbfup{s}}^a\pertmat{D}}) \\
  &\eqtr
  (\pertmat{h}^a + \pertmat{V}^{t,a} + \frac{1}{2}\pertmat{G}^{\gamma, a}(\pertmat{D}) + \pertmat{F}^{\Omega_a}_\mathrm{xc}
  -\frac{\mathrm{i}}{2}\pertmat{T}^a)\pertmat{D}
  + h_\mathrm{nuc}^a + {}^t\tilde{\p}\tilde{\mathbfup{s}}^a\pertmat{D}.
  \end{aligned}
\end{equation}
$\pertmat{F}^{\Omega_a}_\mathrm{xc}$ is the functional derivative matrix
defined in terms of the perturbed overlap distributions
$\tilde{\Omega}^a$.
Further perturbation-strength differentiations and evaluation at zero
perturbation strength will yield the desired response functions:
\begin{subequations}
  \begin{align}
    L^a &\eqavetr
    \mathcal{G}^{00,a} - \mat{S}^a\mat{W} \label{eq:first-order} \\
    L^{ab} &\eqavetr \mathcal{G}^{00,ab}
    + \mathcal{G}^{10,a}\mat{D}^{b}
    + \mathcal{G}^{01,a}\p^{b}
    - \mat{S}^{ab}\mat{W}
    - \mat{S}^{a}\mat{W}^{b} \label{eq:second-order} \\
    L^{abc} &\eqavetr
      \mathcal{G}^{00,abc}
    + \mathcal{G}^{10,ac}\mat{D}^{b}
    + \mathcal{G}^{10,ab}\mat{D}^{c}
    + \mathcal{G}^{20,a}\mat{D}^{b}\mat{D}^{c} \nonumber \\
    &+ \mathcal{G}^{10,a}\mat{D}^{bc}
    + \mathcal{G}^{11,a}\mat{D}^{b}\p^c \label{eq:third-order} \\
    &+ \mathcal{G}^{01,ac}\p^{b}
    + \mathcal{G}^{01,ab}\p^{c}
    + \mathcal{G}^{01,a}\p^{bc}
    + \mathcal{G}^{11,a}\p^{b}\mat{D}^c \nonumber \\
    &- \mat{S}^{abc}\mat{W}
    - \mat{S}^{ab}\mat{W}^{c}
    - \mat{S}^{ac}\mat{W}^{b}
    - \mat{S}^{a}\mat{W}^{bc} \nonumber
  \end{align}
\end{subequations}
and similarly for higher orders.

The governing equations for the perturbed polarization densities are
obtained in complete analogy with the handling of pertubed density
matrices described in \noparcite[ref.][]{Thorvaldsen2008-sg}
Denoting by the superscript $b_N$ a tuple of simulatenous perturbations,
the polarization can be decomposed into frequency components.
Similarly to the treatment of the perturbed density matrices,
we introduce a partition of the frequency components of the
polarization $\p_\omega^{b_N}$ into homogeneous (H) and particular (P)
terms:
\begin{equation}\label{eq:polarization-partition}
  \p_\omega^{b_N} = \p_\mathrm{P}^{b_N} + \p_\mathrm{H}^{b_N} \text{,}
\end{equation}
and we thus get the following system of equations:
\begin{subequations}
\begin{empheq}[left={\empheqlbrace}]{align}
    \V\p_\mathrm{H}^{b_N} &+ \Tr \mathbfup{s}\Dh{b_N} = 0
    \label{eq:homo-polarization-eq}\\
    \V\p_\mathrm{P}^{b_N} &+ \Tr \mathbfup{s}\Dp{b_N}
    = \mathbb{S}^{(n-1)}_\omega \text{.}
 \label{eq:part-polarization-eq}
\end{empheq}
\end{subequations}
$\mathbb{S}^{(n-1)}_\omega$ contains all terms that depend on lower
order density matrices and differentiated source integrals.
The term $\mathbb{S}^{(n-1)}_\omega$ always contains at least a first
derivative of the source integrals and is thus zero if the basis set is
independent of the perturbation tuple under consideration.
Consequently, the particular term of the polarization is nonzero if and
only if the basis set depends on the external perturbation.

Response equations can be obtained by differentiating the
\acrshort{TDSCF} equation with respect to the perturbation tuple
$b_N$:\autocite{Thorvaldsen2008-sg, Ringholm2014-gx}
\begin{equation}\label{eq:response-equation}
 \left[\genHessian - \omega_{b_N}\genMetric \right]\rspParam{b_N} = \RHS{b_N}
\end{equation}
where the generalized Hessian $\genHessian$ and metric $\genMetric$
matrices were introduced and are defined by their transformations on the
response parameters $\rspParam{b_N}$:\autocite{Larsen2000-hj,
Kjaergaard2008-hy}
\begin{subequations}
\begin{align}
  \genHessian\rspParam{b_N} &=
  \mat{G}^\mathrm{KS}([\rspParam{b_N}, \mat{D}]_{\mat{S}})\mat{D}\mat{S}
  -\mat{S}\mat{D}\mat{G}^\mathrm{KS}([\rspParam{b_N},
  \mat{D}]_{\mat{S}}) \nonumber \\
&+\mat{F} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{S}
-\mat{S} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{F}
\label{eq:E2-lintra} \\
&+ \ASCh{b_N}\mat{\esp}\mat{D}\mat{S}
-\mat{S}\mat{D}\mat{\esp}\ASCh{b_N} \nonumber
\\
  \genMetric\rspParam{b_N} &= \mat{S} [\rspParam{b_N}, \mat{D}]_{\mat{S}} \mat{S} \text{.}
  \label{eq:S2-lintra}
\end{align}
\end{subequations}
The generalized Hessian matrix $\genHessian$ includes two types of
classical terms. \emph{Implicit} terms are those included in the
zeroth-order Fock matrix, $\mat{F}$.
The last two terms in Eq. \ref{eq:E2-lintra} are \emph{explicit} terms,
as they explicitly involve the $N$-th order homogenous polarization
variational density.
The nonequilibrium response framework needed in continuum models when
dealing with frequency-dependent properties,\autocite{Cammi1999-rb, Tomasi2005-vm}
will affect the explicit terms in Eq. \ref{eq:E2-lintra} which are
evaluated by means of the optical permittivity $\dield$.
The \acrlong*{RHS} in the response equation only includes terms that
depend on particular contributions up to the desired order or
lower-order perturbed variational densities:\autocite{Ringholm2014-gx}
\begin{equation}\label{eq:RHS}
  \RHS{b_N} =
  \left[\left( \tilde{\mathbfcal{F}}
-
\frac{\mathrm{i}}{2}\pertmat{S}\deriv{}{t}\right)\pertmat{D}\pertmat{S}\right]^{\ominus,
b_N}_\mathrm{P},
\end{equation}
where $\mathbfcal{F}$ is the generalized \acrshort*{KS} matrix, see
\noparcite[ref.][]{Thorvaldsen2008-sg} for details.

A direct solution of the response equation \eqref{eq:response-equation} by inversion
is not feasible due to the size of the $\left[\genHessian -
\omega_{b_N}\genMetric \right]$ matrix. \emph{Subspace iteration}
methods are usually employed, the solution is achieved iteratively
by enlarging a projection subspace of trial vectors.\autocite{Saad2003-oa,
Saad2011-gm, Kauczor2011-rd, Malmqvist2013-vw}
These methods can be implemented efficiently and in matrix-free form
since the linear transformations $\genHessian\rspParam{b_N}$ and
$\genMetric\rspParam{b_N}$ have known forms only involving integral and
(perturbed) density matrices.
Moreover, Jacobi precondinioting of the iterations is usually sufficiently
robust and efficient, due to the diagonal dominance of the generalized
Hessian.
